Predicting I/O-performance in HPC using Artificial Neural Networks

### 2016-06-03 Poster/main.tex
poster Verbesserungen

There is a need for tools that help users to implement efficient input/output (I/O) in their software.
Finding the best access parameters and patterns is difficult, mainly because of the use of complex parallel storage systems in HPC.
%Currently users have to optimize their programs at great expense to each system individually without much assistance.
For the development of tools, to support the implementation of efficient I/O, it is key to have a computational model of the storage system.\medskip

For single hard disk systems such a model can be derived analytically \cite{Ruemmler94anintroduction}, however for the complex storage system of a super computer these models become too difficult to configure \cite{DBLP:conf/npc/ZhangLZJC10}.
Therefore we tried to find a good predictor for I/O-performance using a machine learning approach with artificial neural networks (ANNs).
A hypothesis was then proposed: \textbf{The I/O-path significantly influences the time needed for a file access.}\medskip
%In particular the access time for a file access is more dependent on the I/O-path used in the storage system than on the access size for example.

We used ANNs with different input information for the prediction of access times to test our hypothesis.
For the use of I/O-paths as input information for the ANNs we developed a method to approximate the different I/O-paths the storage system used during a benchmark-test.
This method utilizes error classes.

@@

	ANNs are bio-inspired function approximators.

	%\begin{wrapfigure}{r}{0.6\textwidth}
	%	\begin{center}
	%		\includegraphics[height=1\textwidth]{src/tupel1.png}
	%	\end{center}
	%	\caption{G}
	%	\label{fig:evo}
	%\end{wrapfigure}

	An ANN is able to approximate a function that maps given input vectors on their associated output vectors.
	This is done by adapting the connection weights in the network using gradient decent during a supervised learning process (backpropagation).
	With appropriate topology \textbf{ANNs are able to approximate any continuous functions} on compact subsets of $\mathbb{R}^n$ \cite{cybenko:mcss}.

@@

	The processing of a file access in the storage system can be viewed using the I/O-path, which is \textbf{the path from the invoking processor to the storage medium that contains the data}.
	The resulting access time of the process then depends on the depth of this I/O-path, because storage media further along the path becomes increasingly slower.
	%the further the path leads into the storage system, the slower the storage media gets.
	While the first storage levels (Caches) are very quick to respond, the main memory is already magnitudes slower; the same applies for the step into the parallel storage system that is connected via network to the computer nodes. %\medskip

@@

	Our measurements showed, as can be seen in figure \ref{exploration}, that the access times increase with access sizes.
	However a phenomenon that can be explained with the I/O-path model can be seen as well: \textbf{Access times of measurements with equal parameter values are split into several groups of access times.}
	The used I/O-paths for the file accesses distinguishes these groups of measurements through a step in the magnitude of access time.
	%Different I/O-paths were used for file accesses in these groups, which led to a step in the magnitude of the access time.

	\begin{figure}
		%\flushleft
		\captionsetup[subfigure]{margin={-0.4cm,0cm}}
		\subfloat[Sequential access pattern]{
			%\includegraphics[width=.55\textwidth]{src/plot_SizeSorted_log_read_seq.png}
			\begin{tikzpicture}[spy using outlines={circle,black,magnification=2.2,size=4.5cm, connect spies}]
			\node {\pgfimage[interpolate=true,width=.53\textwidth]{src/plot_SizeSorted_log_read_seq.png}};
			%\draw (3.5,0.5) ellipse (1 and 1);
			\spy on (3.4,0.5) in node [left] at (14,1.25);
			\end{tikzpicture}
		}
		%\hfill
		\\
		\captionsetup[subfigure]{margin={-0.4cm,0cm}}
		\vspace*{0.25cm}	
		\hspace*{-7.365cm}		
		\subfloat[Random access pattern]{
			%\hspace*{-5cm}
			\includegraphics[width=.53\textwidth]{src/plot_SizeSorted_log_read_rnd.png}
		}
		\vspace*{-0.05cm}	
		\caption{Access times of measured file accesses. The access sizes increase from left to right, all measurements with equal parameter values have the same color. In the highlighted section the described step in the access time occurs.}
		\label{exploration}
	\end{figure}

@@

	Different models were used to predict access times of file accesses.
	As baseline model linear regression was used for a simple mapping of access size to access time.
	Additionally three \textbf{models with different input information utilizing ANNs were used}.
	Every ANN-model received information about access sizes, file offsets and access types as input.
	One ANN also received information about the past data throughputs of the system, which can be used to exploit time dependencies of the I/O-performance (see figure \ref{time_dep}). The last ANN received error classes as additional input.

	\begin{figure}
		\subfloat[In this sequence peaks in the access times occur regularly.]{
			\includegraphics[width=.48\textwidth]{src/plot_From100001to100250_read_seq.png}
		}
		\hfill
		\subfloat[Exponential moving average in red to predict peaks.]{
			\includegraphics[width=.48\textwidth]{src/ema2.png}
		}	
		\caption{Time dependencies can be utilized for better access time predictions.}
		\label{time_dep}
	\end{figure}

@@

The idea of error classes is based on the fact that \textbf{the residues of a simple model are characteristic for the groups of measurements with different typical access times} obeserved in figure \ref{exploration}. The residues therefore represent the I/O-paths used by the storage system for the corresponding file accesses. 
%Here a simple model is a model that doesn't discriminate measurements with similar parameters.
Error classes are then obtained by clustering the residues with a k-Means algorithm.

@@

		\begin{figure}
			%\flushleft
			\captionsetup[subfigure]{margin={-0.4cm,0cm}}
			\subfloat[Sequential access pattern]{
				%\includegraphics[width=.55\textwidth]{src/plot_SizeSorted_log_read_seq.png}
				\begin{tikzpicture}[spy using outlines={circle,black,magnification=2.2,size=4cm, connect spies}]
				\node {\pgfimage[interpolate=true,width=.53\textwidth]{src/linreg_error_clustering_seq_all.png}};
				%\draw (3.5,0.5) ellipse (1 and 1);
				\spy on (3.5,0.9) in node [left] at (14,1.25);
				\end{tikzpicture}
			}
			%\hfill
			\\
			\captionsetup[subfigure]{margin={-0.4cm,0cm}}
			\hspace*{-7.365cm}	
			%\vspace*{0.25cm}	
			\subfloat[Random access pattern]{
				%\hspace*{-5cm}
				\includegraphics[width=.53\textwidth]{src/linreg_error_clustering_rnd_all.png}
			}
			\vspace*{-0.1cm}
			\caption{Residues from the linear regression model clustered into error classes. In the highlighted section the differen- tiation of measurements with varying I/O-paths can be seen.}
			\label{error_classes}
		\end{figure}

@@

	From the residues of our models (see table \ref{residues}) it became clear that access times shouldn't be modelled linearly.
	Furthermore the exploitation of time dependencies of the storage system has been found to be difficult.
	As our hypothesis predicted \textbf{it is essential to use knowledge about I/O-paths for precise predictions of access times}.
	The error classes allow the ANN to decrease the error significantly.
	In figure \ref{preds} the benefit of error classes as input can be seen:
	The second ANN is often able to classify file accesses correctly in respect to their I/O-path.

	\begin{table}
		\scriptsize
		\vspace{0.2cm}
		\begin{tabular}{l|r|r|r|r}%
			&  \multicolumn{2}{|c}{MAE (s)}&  \multicolumn{2}{|c}{MSPE (\%)}\\ \hline
			%\toprule
			Modell & seq. access & rnd. access & seq. access & rnd. access\\ \hline
			%\midrule
			Linear regression & $7.6\cdot 10^-5$ & $4.76\cdot 10^-3$ & 59 & 14185  \\
			ANN & $6.0\cdot 10^-5$ & $3.13\cdot 10^-3$ & 22 & 530 \\
			ANN + time dependencies & $5.7\cdot 10^-5$ & $3.05\cdot 10^-3$ & 22 & 619\\
			ANN + error classes & $2.0\cdot 10^-5$ & $1.03\cdot 10^-3$ & 14 & 119\\
			%\bottomrule
		\end{tabular}
		\caption{Mean absolut errors and mean square percentage errors of the models for two different file access patterns.}
		\label{residues}
	\end{table}

@@

	The hypothesis is supported by our data.
	For a good model of HPC storage systems it is therefore necessary to deduce information about I/O-paths.
	If measured access times are available, this can be done with the introduced method.\\
	The method could be a starting point to develop a tool that provides information about I/O-paths used during execution of a program.

>>>>

Tools are demanded that help users of HPC-facilities to implement efficient input/output (I/O) in their programs.
It is difficult to find the best access parameters and patterns, due to complex parallel storage systems.
%Currently users have to optimize their programs at great expense to each system individually without much assistance.
To develop tools, which support the implementation of efficient I/O, a computational model of the storage system is key.\medskip

For single hard disk systems such a model can be derived analytically \cite{Ruemmler94anintroduction}; however, for the complex storage system of a super computer these models become too difficult to configure \cite{DBLP:conf/npc/ZhangLZJC10}.
Therefore we searched for good predictors of I/O-performance using a machine learning approach with artificial neural networks (ANNs).
A hypothesis was then proposed: \textbf{The I/O-path significantly influences the time needed to access a file.}\medskip
%In particular the access time for a file access is more dependent on the I/O-path used in the storage system than on the access size for example.

In our analysis we used ANNs with different input information for the prediction of access times.
To use I/O-paths as input for the ANNs, we developed a method, which approximates the different I/O-paths the storage system used during a benchmark-test.
This method utilizes error classes.

@@

	ANNs are bio-inspired function approximators.
	An ANN is able to approximate a function that maps given input vectors onto their associated output vectors.
	This is done by adapting the connection weights in the network, using gradient decent during a supervised learning process (backpropagation).
	With appropriate topology \textbf{ANNs can approximate any continuous functions} on compact subsets of $\mathbb{R}^n$ \cite{cybenko:mcss}.

@@

	The processing of a file access in the storage system can be viewed using the I/O-path, which is \textbf{the path from the invoking processor to the storage medium that contains the data}.
	The resulting access time of the process depends on the depth of this I/O-path because storage media further along the path are increasingly slower.
	%the further the path leads into the storage system, the slower the storage media gets.
	While the first storage levels (Caches) are very quick to respond, the main memory is already magnitudes slower; the same applies for the step into the parallel storage system, that is connected via network to the computer nodes. %\medskip

@@

	Our measurements have shown, as can be seen in figure \ref{exploration}, that the access times increase with access sizes.
	However, a phenomenon that can be explained with the I/O-path model can be seen as well: \textbf{Access times of measurements with equal parameter values are split into several groups of access times.}
	The file accesses can be distinguished by a step in the magnitude of access time that is caused by the different I/O-paths the system used.
	%Different I/O-paths were used for file accesses in these groups, which led to a step in the magnitude of the access time.

	\begin{figure}
		%\flushleft
		\captionsetup[subfigure]{margin={-0.4cm,0cm}}
		\subfloat[Sequential access pattern]{
			%\includegraphics[width=.55\textwidth]{src/plot_SizeSorted_log_read_seq.png}
			\begin{tikzpicture}[spy using outlines={circle,black,magnification=2.2,size=4.5cm, connect spies}]
			\node {\pgfimage[interpolate=true,width=.53\textwidth]{src/plot_SizeSorted_log_read_seq.png}};
			%\draw (3.5,0.5) ellipse (1 and 1);
			\spy on (3.4,0.5) in node [left] at (14,1.25);
			\end{tikzpicture}
		}
		%\hfill
		\\
		\captionsetup[subfigure]{margin={-0.4cm,0cm}}
		\vspace*{0.25cm}	
		\hspace*{-7.365cm}		
		\subfloat[Random access pattern]{
			%\hspace*{-5cm}
			\includegraphics[width=.53\textwidth]{src/plot_SizeSorted_log_read_rnd.png}
		}
		\vspace*{-0.05cm}	
		\caption{Access times of measured file accesses. The access sizes increase from left to right. All measurements with equal parameter values have the same color. The highlighted area shows a split of measurements with equal parameter values into groups of access times.}
		\label{exploration}
	\end{figure}

@@

	Different models were used to predict access times of file accesses.
	Linear regression was used as a baseline model with a simple mapping of access size to access time.
	Additionally, three \textbf{models with different input information utilizing ANNs were used}.
	Every ANN-model received information about access sizes, file offsets and access types as input.
	One ANN also received information about the past data throughputs of the system, which can be used to exploit time dependencies of the I/O-performance (see figure \ref{time_dep}). The last ANN received error classes as additional input.

	\begin{figure}
		\subfloat[In this sequence, peaks in the access times occur regularly.]{
			\includegraphics[width=.48\textwidth]{src/plot_From100001to100250_read_seq.png}
		}
		\hfill
		\subfloat[Exponential moving average in red to predict peaks.]{
			\includegraphics[width=.48\textwidth]{src/ema2.png}
		}	
		\caption{Time dependencies can be utilized for better access time predictions.}
		\label{time_dep}
	\end{figure}

@@

The idea of error classes is based on the fact that \textbf{the residues of a simple model are characteristic for the groups of measurements with different typical access times} we obeserved in figure \ref{exploration}. The residues therefore represent the I/O-paths used by the storage system for the corresponding file accesses. 
%Here a simple model is a model that doesn't discriminate measurements with similar parameters.
Error classes are obtained by clustering the residues with a k-Means algorithm.

@@

		\begin{figure}
			%\flushleft
			\captionsetup[subfigure]{margin={-0.4cm,0cm}}
			\subfloat[Sequential access pattern]{
				%\includegraphics[width=.55\textwidth]{src/plot_SizeSorted_log_read_seq.png}
				\begin{tikzpicture}[spy using outlines={circle,black,magnification=2.2,size=4cm, connect spies}]
				\node {\pgfimage[interpolate=true,width=.53\textwidth]{src/linreg_error_clustering_seq_all.png}};
				%\draw (3.5,0.5) ellipse (1 and 1);
				\spy on (3.5,0.9) in node [left] at (14,1.25);
				\end{tikzpicture}
			}
			%\hfill
			\\
			\captionsetup[subfigure]{margin={-0.4cm,0cm}}
			\hspace*{-7.365cm}	
			%\vspace*{0.25cm}	
			\subfloat[Random access pattern]{
				%\hspace*{-5cm}
				\includegraphics[width=.53\textwidth]{src/linreg_error_clustering_rnd_all.png}
			}
			\vspace*{-0.1cm}
			\caption{Residues from the linear regression model clustered into error classes. Measurements with equal parameter values can be differentiated through error classes.}
			\label{error_classes}
		\end{figure}

@@

	From the residues of our models (see table \ref{residues}) it became clear that access times shouldn't be modelled linearly.
	Furthermore, it has been found to be difficult to exploit time dependencies of the storage system.
	As our hypothesis predicted, \textbf{it is essential to use knowledge about I/O-paths for precise predictions of access times}.
	The error classes allow the ANN to decrease the error significantly.
	In figure \ref{preds} the benefit of error classes as input can be seen:
	The second ANN is often able to classify file accesses correctly in respect to their I/O-path.

	\begin{table}
		\scriptsize
		\vspace{0.2cm}
		\begin{tabular}{l|r|r|r|r}%
			&  \multicolumn{2}{|c}{MAE (s)}&  \multicolumn{2}{|c}{MSPE (\%)}\\ \hline
			%\toprule
			Modell & seq. access & rnd. access & seq. access & rnd. access\\ \hline
			%\midrule
			Linear regression & $7.6\cdot 10^-5$ & $4.76\cdot 10^-3$ & 59 & 14185  \\
			ANN & $6.0\cdot 10^-5$ & $3.13\cdot 10^-3$ & 22 & 530 \\
			ANN + time dependencies & $5.7\cdot 10^-5$ & $3.05\cdot 10^-3$ & 22 & 619\\
			ANN + error classes & $2.0\cdot 10^-5$ & $1.03\cdot 10^-3$ & 14 & 119\\
			%\bottomrule
		\end{tabular}
		\caption{Mean absolut errors and mean square percentage errors of the models for the two different file access patterns.}
		\label{residues}
	\end{table}

@@

	The hypothesis is supported by our data.
	It is therefore necessary to deduce information about I/O-paths to build a good model of HPC storage systems.
	If measured access times are available, this can be done with the introduced method.\\
	The method could be a starting point to develop a tool that provides information about I/O-paths used during execution of a program.
